### 2023-07-06 paper/localpsf.tex
edits with Noemi. Paper should be done

% REQUIRED
\begin{abstract}
	We present an efficient matrix-free point spread function (PSF) method for approximating operators that have locally supported non-negative integral kernels. The method computes impulse responses of the operator at scattered points, and interpolates these impulse responses to approximate entries of the integral kernel. To compute impulse responses efficiently, we apply the operator to Dirac combs associated with batches of point sources, which are chosen by solving an ellipsoid packing problem. The ability to rapidly evaluate kernel entries allows us to construct a hierarchical matrix (H-matrix) approximation of the operator. Further matrix computations are then performed with fast H-matrix methods. We demonstrate the effectiveness of this PSF-based method by using it to build preconditioners for the Hessian operator arising in two inverse problems governed by partial differential equations (PDEs): inversion for the basal friction coefficient in an ice sheet flow problem and for the initial condition in a diffusive transport problem. While for many ill-posed inverse problems the Hessian of the data misfit term exhibits a low rank structure, and hence a low rank approximation is suitable, for many problems of practical interest the numerical rank of the Hessian is still large. The Hessian impulse responses on the other hand typically become more local as the numerical rank increases, which benefits the PSF-based method. Numerical results reveal that the PSF-based preconditioner clusters the spectrum of the preconditioned Hessian near one, yielding roughly $5\times$--$10\times$ reductions in the required number of PDE solves, as compared to classical regularization-based preconditioning and no preconditioning. We also present a comprehensive numerical study for the influence of various parameters (that control the shape of the impulse responses and the rank of the Hessian) on the effectiveness of the approximation. The results show that the PSF-based preconditioners are able to form good approximations of high-rank Hessians using only a small number of operator applications.
\end{abstract}

% REQUIRED
\begin{keywords}
 	data scalability, Hessian, hierarchical matrix, high-rank, impulse response, local translation invariance, matrix-free, moment methods, operator approximation, PDE constrained inverse problems, point spread function, product convolution  
\end{keywords}

@@

\begin{figure}
	\begin{subfigure}{0.49\textwidth}
		\centering
		\includegraphics[scale=0.3]{impulse_batch1.png}
		\caption{One impulse response batch}
		\label{fig:batches_intro}
	\end{subfigure}
	\begin{subfigure}{0.49\textwidth}
		\centering
		\includegraphics[scale=0.7]{mean_displacement_invariance.pdf}
		\caption{Local mean displacement invariance}
		\label{fig:mean_displacement_invariance}
	\end{subfigure}
%	\begin{subfigure}{0.32\textwidth}
%		\centering
%		\includegraphics[scale=0.34]{impulse_batch3.png}
%	\end{subfigure}
	\caption{(\ref{fig:batches_intro}) One batch, $\eta_b$, of normalized impulse responses, $\impulseresponse_{x}$, that arise from applying $\Aop$ to a weighted sum of scattered point sources (see Section \ref{sec:get_impulse_response}). Here $\Aop$ is the ice sheet inverse problem data misfit Gauss-Newton Hessian described in Section~\ref{sec:numerical_results}. 
	Black stars are point source locations. Shading shows the magnitude of the normalized impulse responses (darker means larger function values). Dashed gray ellipses are estimated impulse response support ellipsoids based on the moment method in Section \ref{sec:intromoments}. The large circle is $\partial \Omega$. (\ref{fig:mean_displacement_invariance}) Illustration of impulse responses, $\impulseresponse_x$ and $\impulseresponse_{x'}$, that arise from applying $\Aop$ to point sources at two points, $x$ and $x'$. The operator $\Aop$ is locally mean displacement invariant (Section \ref{sec:local_mean_displacement_invariance}) if $\impulseresponse_{x}(y) \approx \impulseresponse_{x'}\left(y - \spatialmean(x) + \spatialmean(x')\right)$ when $x$ is close to $x'$. Here $\spatialmean(z)$ denotes the mean (center of mass) of $\phi_z$.
	}
	\label{fig:batches_and_mdi}
\end{figure}

We use \emph{impulse response interpolation} to form a high rank approximation of $\Aop$ using a small number of operator applications. The impulse response, $\phi_x$, associated with a point, $x$, is the Riesz representation\footnote{Recall that the Riesz representative of a functional $\rho \in L^2(\Omega)'$ with respect to the $L^2$ inner product is the unique function $\rho^* \in L^2(\Omega)$ such that $\rho(w) = \left(\rho^*,w\right)_{L^2(\Omega)}$ for all $w \in L^2(\Omega)$.} of the linear functional that results from applying $\Aop$ to a delta distribution (i.e., point source, impulse) centered at $x$. We compute batches of impulse responses by applying $\Aop$ to weighted sums of delta distributions associated with batches of points scattered throughout the domain (see Figure~\ref{fig:batches_intro}). Then we interpolate translated and scaled versions of these impulse responses to approximate entries of the operator's integral kernel. 
Picking the batches of points requires us to estimate the supports of the impulse responses $\phi_x$ \emph{before} they are computed. Ellipsoid estimates for the supports of all $\phi_x$ are determined a-priori via a moment method that involves applying $\Aop^T$ to a small number of polynomial functions (see Section~\ref{sec:intromoments}). We are inspired by resolution analysis in seismic imaging, in which
$\Aop^T$ is applied to a random noise function, and the width of the support $\phi_x$ is estimated to be the autocorrelation length of the resultant function near $x$~\cite{FichtnerLeeuwen15,TrampertFichtnerRitsema13}. The moment method that we use estimates the support of $\phi_x$ more accurately than random noise probing in resolution analysis, at the cost of the additional constraint that $\Aop$ has a non-negative integral kernel.

@@

In distributed parameter inverse problems governed by PDEs, one seeks to infer an unknown spatially varying parameter field from limited observations of a state variable that depends on the parameter implicitly through the solution of a PDE. 
Conventionally, the inverse problem is formulated using either a deterministic framework~\cite{AsterBorchersThurber18,BanksKunisch89,Vogel02},
or a Bayesian probabilistic framework~\cite{KaipioSomersalo05,Stuart10a,Tarantola05}. In the deterministic framework, one solves an optimization problem to find the parameter that best fits the observations, subject to appropriate regularization~\cite{EnglHankeNeubauer96,Vogel02}. In the probabilistic framework, Bayes' theorem combines the observations with prior information to form a posterior distribution over the space of all possible parameter fields, and computations are performed to extract statistical information about the parameter from this posterior. The Hessian of the objective function with respect to the parameter in the determinstic optimization problem and the Hessian of the negative log posterior in the Bayesian setting are equal or approximately equal under typical noise, regularization, and prior models, so we refer to both of these Hessians as ``the Hessian.'' The Hessian consists of a data misfit term (the \emph{data misfit Hessian}), which depends on a discrepancy between the observations and the associated model predictions, and a regularization or prior term (the \emph{regularization Hessian}) which does not depend on the observations. For more details on the Hessian, see~\cite{Alger19,GhattasWillcox21,VillaPetraGhattas21}.

@@

Owing to the implicit dependence of predicted observations on the parameter, entries of the Hessian are not easily accessible. Rather, the Hessian may be applied to a vector via a computational process that involves solving a pair of forward and adjoint PDEs which are linearizations of the original PDE~\cite{GhattasWillcox21}
%,Gunzburger03}
. The most popular matrix-free Hessian approximation methods are based on low rank approximation of either the data misfit Hessian, or the data misfit Hessian preconditioned by the regularization Hessian, e.g.,~\cite{BuiEtAl13,CuiEtAl14,FlathEtAl11,PetraEtAl14,SpantiniEtAl15}. Krylov methods such as Lanczos or randomized methods~\cite{Cheng05,HalkoMartinssonTropp11} are typically used to construct these low rank approximations by applying the Hessian to vectors. Using these methods, the required number of Hessian applications (and hence the required number of PDE solves) is proportional to the rank of the low rank approximation. Low rank approximation methods are justified by arguing that the numerical rank of the data misfit Hessian is insensitive to the dimension of the discretized parameter. This means that the required number of PDE solves remains the same as the mesh used to discretize the parameter is refined.
However, in many inverse problems of practical interest the numerical rank of the data misfit Hessian, while mesh independent, is still large, which makes it costly to approximate the Hessian using low rank approximation methods~\cite{AmbartsumyanEtAl20,BuiEtAl13,IsaacEtAl15}.

@@

We focus on approximating operators that satisfy the following conditions: 
\begin{enumerate}
	\item The kernel $\Aker$ is sufficiently regular so that $\phi_x$ is well-defined for all $x\in \Omega$.
	\item The supports of the impulse responses $\impulseresponse_x$ are contained in localized regions.
	\item The integral kernel is non-negative in the sense that $$\Aker(y,x) \ge 0$$ for all $(y,x) \in \Omega \times \Omega$.\footnote{Note that having a non-negative integral kernel is different from positive semi-definiteness. The operator $\Aop$ need not be positive semi-definite to use the proposed method, and positive semi-definite operators need not have a non-negative integral kernel.}
\end{enumerate}
The proposed method may still perform well if these conditions are relaxed slightly. It is acceptable if the support of $\impulseresponse_x$ is not perfectly contained in a localized region, so long as the bulk of the ``mass'' of $\impulseresponse_x$ is contained in a localized region. The integral kernel does not need to be non-negative for all pairs of points $(y,x) \in \Omega \times \Omega$ so long as it is non-negative for the vast majority of pairs of points $(y,x)$, and so long as the negative numbers are comparatively small. If these conditions are violated, the method will incur additional error, and depending on the severity of the violation, the proposed method may fail.

@@

We use the proposed method to approximate the Newton (or Gauss-Newton) Hessians
in inverse problems governed by PDEs which model steady state ice
sheet flow~\cite{PetraEtAl12} (Section~\ref{sec:ice}) and transport
and flow of a contaminant~\cite{PetraStadler11}
(Section~\ref{sec:adv}). These inverse problems are described in detail in their respective sections.
In both cases, to reconstruct the unknown parameter fields, denoted $q$, the inverse problems are formulated as nonlinear least squares
optimization problems, whose objective functions consist of a data misfit term
(between the observations and model output) and a bi-Laplacian
regularization term following~\cite{VillaPetraGhattas21}. The
regularization is centered at a constant function $q_0(x)$. To
mitigate boundary effects
%For stokes q_0(x) = 10.5$%
we use a constant coefficient Robin boundary condition as
in~\cite{Roininen14}. The parameters for the bi-Laplacian operator are chosen
so that the Green's function of the Hessian of the regularization has
a characteristic length of $0.25$ of the domain radius. For the
specific setup, we refer the reader to~\cite[Section
  2.2]{VillaPetraGhattas21}. In all numerical results we choose the
regularization parameter (which controls the overall strength of the regularization) using the Morozov discrepancy
principle~\cite{Vogel02}.

We solve the ice sheet inverse problem with an inexact Newton preconditioned
conjugate gradient (PCG) scheme and a globalizing Armijo line search~\cite{NocedalWright99}. The Newton updates,
$\mathbf{\searchdir}$, are obtained by solving 
%
\begin{equation}
	\label{eq:newton_system}
	\mathbf{H} \mathbf{\searchdir} = - \mathbf{g} \qquad \text{or} \qquad \mathbf{H}_\text{gn} \mathbf{\searchdir} = - \mathbf{g},
\end{equation}
%
wherein we choose the initial guess as the discretization of the constant
function $q_0$. Here $\mathbf{g}$, $\mathbf{H}$ and $\mathbf{H}_\text{gn}$ are
the discretized gradient, Hessian, and Gauss-Newton Hessian of the
inverse problem objective function, respectively, evaluated at the current Newton iterate.
To ensure positive definiteness of the Hessian we use
$\mathbf{H}_\text{gn}$ for the first five iterations, and $\mathbf{H}$ for all
subsequent iterations. The Newton iterations are terminated when
$\|\mathbf{g}\| < 10^{-6} \|\mathbf{g}_0\|$, where $\mathbf{g}_0$ is the gradient evaluated at the initial guess. Systems~\eqref{eq:newton_system} are solved inexactly using an inner PCG
iteration, which is terminated early based on the Eisenstat-Walker \cite{EisenstatWalker96} and Steihaug \cite{Steihaug83} conditions. The inverse problem governed by the advection-diffusion PDE is linear, hence Newton's method converges in one iteration. In this case the Newton linear system,~\eqref{eq:newton_system}, is solved using PCG, using termination tolerances described in Section \ref{sec:adv}.

@@

For this example, we consider a sheet of ice flowing down a mountain (see Figure~\ref{fig:ice_mountain_mesh}). Given observations of the tangential component of the ice velocity on the top surface of the ice, we invert for the logarithm of the unknown spatially varying basal friction Robin coefficient field, which governs the resistance to sliding along the base of the ice sheet. 
The setup, which we briefly summarize, follows~\cite{IsaacEtAl15,PetraEtAl12}.
The region of ice is denoted by
$\icedomain \subset\mathbb{R}^{3}$. The basal, lateral and top parts of the boundary $\partial \icedomain$ are denoted by $\Gamma_{b}$, $\Gamma_{l}$, and $\Gamma_{t}$, respectively.
The governing equations are the linear incompressible Stokes
equations,
%
%
\begin{subequations}\label{Stokeseqn}
	\begin{align}
	%-\nabla \cdot \stress(v, p)
	%&=\bodyforce \,\,\,\,\,\text{ in }\icedomain, \label{Stokeseqn:1}\\
	%\nabla\cdot \velocity
	  %&=0\,\,\,\,\,\,\text{ in }\icedomain, \label{Stokeseqn:2}\\
          -\nabla \cdot \stress(v, p)=\bodyforce \text{ and }
          \nabla\cdot \velocity &=0 \quad \text{ in }\icedomain, \label{Stokeseqn:1}\\
	\stress(v, p)\normalvec
	&=0\,\,\,\,\,\,\text{ on }\Gamma_{t}, \label{Stokeseqn:3}\\
	\velocity \cdot\normalvec =0 \text{ and } \tangentop\left(\stress(v, p)\normalvec+\exp\left(\basalfriction\right)\velocity\right)
	&=0\,\,\,\,\,\,\text{ on }\Gamma_{b},\label{Stokeseqn:4} \\
	\stress(v, p)\normalvec+\stokesrobincoeff \velocity &=0\,\,\,\,\,\,\text{ on }\Gamma_{l}.\label{Stokeseqn:5}	
	\end{align}
\end{subequations}
%
%
The solution to these equations is the pair $(\velocity, \pressure)$, where $\velocity$ is the ice flow velocity field\footnote{We do not use bold to denote vector or tensor fields to avoid confusion with vectors that arise from finite element discretizations, which are already denoted with bold.}, and $\pressure$ is the pressure field. Here, $\basalfriction$ is the unknown logarithmic
basal friction field (large $\basalfriction$ corresponds to large resistance to sliding) defined on the surface $\Gamma_b$. The quantity $\bodyforce$ is the body force density due to gravity, $\stokesrobincoeff=10^6$ is a Robin boundary condition constant,  $\normalvec$ is the outward unit normal and $\tangentop$ is the tangential projection operator that restricts a vector field to its tangential component along the boundary. 
We employ a Newtonian constitutive law, $\stress(v, p)= 2\eta
\dot{\strain}(v) -I\pressure$,
where $\stress$ is the stress tensor and
$\dot{\strain}(v)= \frac{1}{2}\left(
\nabla\velocity+\nabla\velocity^{\top} \right)$ is the strain rate tensor~\cite{Glen1955}. Here $\eta$ is the viscosity and $I$ is the identity operator. 
%Glen's exponent has been chosen to be one, which provides for a linear Stokes model. 
Note that while the PDE is linear, the parameter-to-solution map, $\basalfriction \mapsto (\velocity, \pressure)$, is nonlinear.

The pressure, $\pressure$, is discretized with first order scalar continuous Galerkin finite elements defined on a mesh of tetrahedra. The velocity, $\velocity$, is discretized with second order continuous Galerkin finite elements on the same mesh. The parameter $\basalfriction$ is discretized with first order scalar continuous Galerkin finite elements on the mesh of triangles that results from restricting the tetrahedral mesh to the basal boundary, $\Gamma_b$. Note that $\Gamma_b$ is a two-dimensional surface embedded in three dimensions due to the mountain topography. We also generate a flattened version of $\Gamma_b$, denoted by $\Omega \subset \mathbb{R}^2$, by ignoring the height coordinate\footnote{The proposed PSF method requires translating impulse responses. This requires either a flat domain, or a notion of local parallel transport.}. The parameter $\basalfriction$ is viewed as a function on $\Gamma_b$ for the purpose of solving the Stokes equations, and as a function on $\Omega$ for the purpose of building Hessian approximations and defining the regularization. 
%This flattening is necessary because our operator approximation method requires adding points on this surface together to get other points on the surface, and this is only well-defined in flat space. 
The observations are generated by adding multiplicative
Gaussian noise to the tangential component of the velocity field
restricted to the top surface of the geometry. We use $5\%$ noise in all cases, except for Figure~\ref{fig:stokes_reconstructions} and Table \ref{tab:condition_number} where the noise is varied from $1\%$ to $25\%$ and the regularization is determined by the Morozov discrepancy principle for each noise level. The true basal friction
coefficient and resulting velocity fields, which are obtained by
solving~\eqref{Stokeseqn}, are shown in Figure~\ref{fig:true_beta_u}.

@@

\begin{figure}
	\begin{subfigure}{0.32\textwidth}
		\includegraphics[scale=0.17]{mstar_0.25noise_cropped.png}
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
		\includegraphics[scale=0.17]{mstar_0.05noise_cropped.png}
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
	\includegraphics[scale=0.17]{mstar_0.01noise_cropped.png}
    \end{subfigure}
	\caption{(Ice sheet) The log basal friction parameter, with color scale as in Figure~\ref{fig:true_beta}, computed from the PDE-constrained optimization problem with noise levels: $25\%$ (left), $5.0\%$ (middle), and $1.0\%$ (right).}
	\label{fig:stokes_reconstructions} 
\end{figure}

>>>>

% REQUIRED
\begin{abstract}
	We present an efficient matrix-free point spread function (PSF) method for approximating operators that have locally supported non-negative integral kernels. The method computes impulse responses of the operator at scattered points, and interpolates these impulse responses to approximate entries of the integral kernel. To compute impulse responses efficiently, we apply the operator to Dirac combs associated with batches of point sources, which are chosen by solving an ellipsoid packing problem. The ability to rapidly evaluate kernel entries allows us to construct a hierarchical matrix (H-matrix) approximation of the operator. Further matrix computations are then performed with fast H-matrix methods. We demonstrate the effectiveness of this PSF-based method by using it to build preconditioners for the Hessian operator arising in two inverse problems governed by partial differential equations (PDEs): inversion for the basal friction coefficient in an ice sheet flow problem and for the initial condition in an advective-diffusive transport problem. While for many ill-posed inverse problems the Hessian of the data misfit term exhibits a low rank structure, and hence a low rank approximation is suitable, for many problems of practical interest the numerical rank of the Hessian is still large. The Hessian impulse responses on the other hand typically become more local as the numerical rank increases, which benefits the PSF-based method. Numerical results reveal that the PSF-based preconditioner clusters the spectrum of the preconditioned Hessian near one, yielding roughly $5\times$--$10\times$ reductions in the required number of PDE solves, as compared to classical regularization-based preconditioning and no preconditioning. We also present a comprehensive numerical study for the influence of various parameters (that control the shape of the impulse responses and the rank of the Hessian) on the effectiveness of the approximation. The results show that the PSF-based preconditioners are able to form good approximations of high-rank Hessians using only a small number of operator applications.
\end{abstract}

% REQUIRED
\begin{keywords}
 	data scalability, Hessian, hierarchical matrix, high-rank, impulse response, local translation invariance, matrix-free, moment methods, operator approximation, PDE constrained inverse problems, point spread function, preconditioning, product convolution
\end{keywords}

@@

\begin{figure}
	\begin{subfigure}[b]{0.49\textwidth}
		\centering
		\includegraphics[scale=0.3]{impulse_batch1.png}
		\caption{One impulse response batch}
		\label{fig:batches_intro}
	\end{subfigure}
	\begin{subfigure}[b]{0.49\textwidth}
		\centering
		\includegraphics[scale=0.7]{mean_displacement_invariance.pdf}
		\caption{Local mean displacement invariance}
		\label{fig:mean_displacement_invariance}
	\end{subfigure}
%	\begin{subfigure}{0.32\textwidth}
%		\centering
%		\includegraphics[scale=0.34]{impulse_batch3.png}
%	\end{subfigure}
	\caption{(\ref{fig:batches_intro}) One batch, $\eta_b$, of normalized impulse responses, $\impulseresponse_{x}$, that arise from applying $\Aop$ to a weighted sum of scattered point sources (see Section \ref{sec:get_impulse_response}). Here $\Aop$ is the ice sheet inverse problem data misfit Gauss-Newton Hessian described in Section~\ref{sec:numerical_results}. 
	Black stars are point source locations. Shading shows the magnitude of the normalized impulse responses (darker means larger function values). Dashed gray ellipses are estimated impulse response support ellipsoids based on the moment method in Section \ref{sec:intromoments}. The large circle is $\partial \Omega$. (\ref{fig:mean_displacement_invariance}) Illustration of impulse responses, $\impulseresponse_x$ and $\impulseresponse_{x'}$, corresponding to points $x$ and $x'$. The operator $\Aop$ is locally mean displacement invariant (Section \ref{sec:local_mean_displacement_invariance}) if $\impulseresponse_{x}(y) \approx \impulseresponse_{x'}\left(y - \spatialmean(x) + \spatialmean(x')\right)$ when $x$ is close to $x'$. Here $\spatialmean(z)$ denotes the mean (center of mass) of $\phi_z$.
	}
	\label{fig:batches_and_mdi}
\end{figure}

We use \emph{impulse response interpolation} to form a high rank approximation of $\Aop$ using a small number of operator applications. The impulse response, $\phi_x$, associated with a point, $x$, is the Riesz representation\footnote{Recall that the Riesz representative of a functional $\rho \in L^2(\Omega)'$ with respect to the $L^2$ inner product is the unique function $\rho^* \in L^2(\Omega)$ such that $\rho(w) = \left(\rho^*,w\right)_{L^2(\Omega)}$ for all $w \in L^2(\Omega)$.} of the linear functional that results from applying $\Aop$ to a delta distribution (i.e., point source, impulse) centered at $x$. We compute batches of impulse responses by applying $\Aop$ to weighted sums of delta distributions associated with batches of points scattered throughout the domain (see Figure~\ref{fig:batches_intro}). Then we interpolate translated and scaled versions of these impulse responses to approximate entries of the operator's integral kernel. 
Picking the batches of points requires us to estimate the supports of the impulse responses $\phi_x$ \emph{before} they are computed. Ellipsoid estimates for the supports of all $\phi_x$ are determined a-priori via a moment method that involves applying $\Aop^T$ to a small number of polynomial functions (see Section~\ref{sec:intromoments}). We are inspired by resolution analysis in seismic imaging, in which
$\Aop^T$ is applied to a random noise function, and the width of the support $\phi_x$ is estimated to be the autocorrelation length of the resultant function near $x$~\cite{FichtnerLeeuwen15,TrampertFichtnerRitsema13}. The moment method that we use estimates the support of $\phi_x$ more accurately than random noise probing in resolution analysis, at the cost of the additional constraint that $\Aop$ has a non-negative integral kernel. Adding more batches yields impulse responses at more points, increasing the accuracy of the approximation at the cost of one operator application per batch.

@@

In distributed parameter inverse problems governed by PDEs, one seeks to infer an unknown spatially varying parameter field from limited observations of a state variable that depends on the parameter implicitly through the solution of a PDE. 
Conventionally, the inverse problem is formulated using either a deterministic framework~\cite{BanksKunisch89,Vogel02},
or a Bayesian probabilistic framework~\cite{KaipioSomersalo05,Stuart10a,Tarantola05}. In the deterministic framework, one solves an optimization problem to find the parameter that best fits the observations, subject to appropriate regularization~\cite{EnglHankeNeubauer96,Vogel02}. In the probabilistic framework, Bayes' theorem combines the observations with prior information to form a posterior distribution over the space of all possible parameter fields, and computations are performed to extract statistical information about the parameter from this posterior. The Hessian of the objective function with respect to the parameter in the determinstic optimization problem and the Hessian of the negative log posterior in the Bayesian setting are equal or approximately equal under typical noise, regularization, and prior models, so we refer to both of these Hessians as ``the Hessian.'' The Hessian consists of a data misfit term (the \emph{data misfit Hessian}), which depends on a discrepancy between the observations and the associated model predictions, and a regularization or prior term (the \emph{regularization Hessian}) which does not depend on the observations. For more details on the Hessian, see~\cite{Alger19,GhattasWillcox21,VillaPetraGhattas21}.

@@

Owing to the implicit dependence of predicted observations on the parameter, entries of the Hessian are not easily accessible. Rather, the Hessian may be applied to a vector via a computational process that involves solving a pair of forward and adjoint PDEs which are linearizations of the original PDE~\cite{GhattasWillcox21,PetraStadler11}.
%,Gunzburger03}
The most popular matrix-free Hessian approximation methods are based on low rank approximation of either the data misfit Hessian, or the data misfit Hessian preconditioned by the regularization Hessian, e.g.,~\cite{BuiEtAl13,CuiEtAl14,FlathEtAl11,PetraEtAl14,SpantiniEtAl15}. Krylov methods such as Lanczos or randomized methods~\cite{Cheng05,HalkoMartinssonTropp11} are typically used to construct these low rank approximations by applying the Hessian to vectors. Using these methods, the required number of Hessian applications (and hence the required number of PDE solves) is proportional to the rank of the low rank approximation. Low rank approximation methods are justified by arguing that the numerical rank of the data misfit Hessian is insensitive to the dimension of the discretized parameter. This means that the required number of PDE solves remains the same as the mesh used to discretize the parameter is refined.
However, in many inverse problems of practical interest the numerical rank of the data misfit Hessian, while mesh independent, is still large, which makes it costly to approximate the Hessian using low rank approximation methods~\cite{AmbartsumyanEtAl20,BuiEtAl13,IsaacEtAl15}.

@@

We focus on approximating operators that satisfy the following conditions: 
\begin{enumerate}
	\item The kernel $\Aker$ is sufficiently regular so that $\phi_x$ is well-defined for all $x\in \Omega$.
	\item The supports of the impulse responses $\impulseresponse_x$ are contained in localized regions.
	\item The integral kernel is non-negative\footnote{Note that having a non-negative integral kernel is different from positive semi-definiteness. The operator $\Aop$ need not be positive semi-definite to use the proposed method, and positive semi-definite operators need not have a non-negative integral kernel.} in the sense that $$\Aker(y,x) \ge 0 \quad \text{for all} \quad (y,x) \in \Omega \times \Omega.$$
\end{enumerate}
The proposed method may still perform well if these conditions are relaxed slightly. It is acceptable if the support of $\impulseresponse_x$ is not perfectly contained in a localized region, so long as the bulk of the ``mass'' of $\impulseresponse_x$ is contained in a localized region. The integral kernel does not need to be non-negative for all pairs of points $(y,x) \in \Omega \times \Omega$ so long as it is non-negative for the vast majority of pairs of points $(y,x)$, and so long as the negative numbers are comparatively small. If these conditions are violated, the method will incur additional error, and depending on the severity of the violation, the proposed method may fail.

@@

We use the proposed method to approximate the Newton (or Gauss-Newton) Hessians
in inverse problems governed by PDEs which model steady state ice
sheet flow~\cite{PetraEtAl12} (Section~\ref{sec:ice}) and advective-diffusive transport of a contaminant~\cite{PetraStadler11}
(Section~\ref{sec:adv}). These inverse problems are described in detail in their respective sections.
In both cases, to reconstruct the unknown parameter fields, denoted $q$, the inverse problems are formulated as nonlinear least squares
optimization problems, whose objective functions consist of a data misfit term
(between the observations and model output) and a bi-Laplacian
regularization term following~\cite{VillaPetraGhattas21}. The
regularization is centered at a constant function $q_0(x)$. To
mitigate boundary effects
%For stokes q_0(x) = 10.5$%
we use a constant coefficient Robin boundary condition as
in~\cite{Roininen14}. The parameters for the bi-Laplacian operator are chosen
so that the Green's function of the Hessian of the regularization has
a characteristic length of $0.25$ of the domain radius. For the
specific setup, we refer the reader to~\cite[Section
  2.2]{VillaPetraGhattas21}. In all numerical results we choose the
regularization parameter (which controls the overall strength of the regularization) using the Morozov discrepancy
principle~\cite{Vogel02}.

We solve the ice sheet inverse problem with an inexact Newton preconditioned
conjugate gradient (PCG) scheme and a globalizing Armijo line search~\cite{NocedalWright99}. The Newton search directions,
$\mathbf{\searchdir}$, are obtained by solving 
%
\begin{equation}
	\label{eq:newton_system}
	\mathbf{H} \mathbf{\searchdir} = - \mathbf{g} \qquad \text{or} \qquad \mathbf{H}_\text{gn} \mathbf{\searchdir} = - \mathbf{g},
\end{equation}
%
wherein we choose the initial guess as the discretization of the constant
function $q_0$. Here $\mathbf{g}$, $\mathbf{H}$ and $\mathbf{H}_\text{gn}$ are
the discretized gradient, Hessian, and Gauss-Newton Hessian of the
inverse problem objective function, respectively, evaluated at the current Newton iterate.
To ensure positive definiteness of the Hessian we use
$\mathbf{H}_\text{gn}$ for the first five iterations, and $\mathbf{H}$ for all
subsequent iterations. The Newton iterations are terminated when
$\|\mathbf{g}\| < 10^{-6} \|\mathbf{g}_0\|$, where $\mathbf{g}_0$ is the gradient evaluated at the initial guess. Systems~\eqref{eq:newton_system} are solved inexactly using an inner PCG
iteration, which is terminated early based on the Eisenstat-Walker \cite{EisenstatWalker96} and Steihaug \cite{Steihaug83} conditions. The inverse problem governed by the advection-diffusion PDE is linear, hence Newton's method converges in one iteration. In this case the Newton linear system,~\eqref{eq:newton_system}, is solved using PCG, using termination tolerances described in Section \ref{sec:adv}.

@@

For this example, we consider a sheet of ice flowing down a mountain (see Figure~\ref{fig:ice_mountain_mesh}). Given observations of the tangential component of the ice velocity on the top surface of the ice, we invert for the logarithm of the unknown spatially varying basal friction Robin coefficient field, which governs the resistance to sliding along the base of the ice sheet. 
The setup, which we briefly summarize, follows~\cite{IsaacEtAl15,PetraEtAl12}.
The region of ice is denoted by
$\icedomain \subset\mathbb{R}^{3}$. The basal, lateral and top parts of the boundary $\partial \icedomain$ are denoted by $\Gamma_{b}$, $\Gamma_{l}$, and $\Gamma_{t}$, respectively.
The governing equations are the linear incompressible Stokes
equations,
%
%
\begin{subequations}\label{Stokeseqn}
	\begin{align}
	%-\nabla \cdot \stress(v, p)
	%&=\bodyforce \,\,\,\,\,\text{ in }\icedomain, \label{Stokeseqn:1}\\
	%\nabla\cdot \velocity
	  %&=0\,\,\,\,\,\,\text{ in }\icedomain, \label{Stokeseqn:2}\\
          -\nabla \cdot \stress(v, p)=\bodyforce \text{ and }
          \nabla\cdot \velocity &=0 \quad \text{ in }\icedomain, \label{Stokeseqn:1}\\
	\stress(v, p)\normalvec
	&=0\,\,\,\,\,\,\text{ on }\Gamma_{t}, \label{Stokeseqn:3}\\
	\velocity \cdot\normalvec =0 \text{ and } \tangentop\left(\stress(v, p)\normalvec+\exp\left(\basalfriction\right)\velocity\right)
	&=0\,\,\,\,\,\,\text{ on }\Gamma_{b},\label{Stokeseqn:4} \\
	\stress(v, p)\normalvec+\stokesrobincoeff \velocity &=0\,\,\,\,\,\,\text{ on }\Gamma_{l}.\label{Stokeseqn:5}	
	\end{align}
\end{subequations}
%
%
The solution to these equations is the pair $(\velocity, \pressure)$, where $\velocity$ is the ice flow velocity field\footnote{We do not use bold to denote vector or tensor fields to avoid confusion with vectors that arise from finite element discretizations, which are already denoted with bold.} and $\pressure$ is the pressure field. Here, $\basalfriction$ is the unknown logarithmic
basal friction field (large $\basalfriction$ corresponds to large resistance to sliding) defined on the surface $\Gamma_b$. The quantity $\bodyforce$ is the body force density due to gravity, $\stokesrobincoeff=10^6$ is a Robin boundary condition constant,  $\normalvec$ is the outward unit normal and $\tangentop$ is the tangential projection operator that restricts a vector field to its tangential component along the boundary. 
We employ a Newtonian constitutive law, $\stress(v, p)= 2\eta
\dot{\strain}(v) -I\pressure$,
where $\stress$ is the stress tensor and
$\dot{\strain}(v)= \frac{1}{2}\left(
\nabla\velocity+\nabla\velocity^{\top} \right)$ is the strain rate tensor~\cite{IsaacEtAl15}. Here $\eta$ is the viscosity and $I$ is the identity operator. 
%Glen's exponent has been chosen to be one, which provides for a linear Stokes model. 
Note that while the PDE is linear, the parameter-to-solution map, $\basalfriction \mapsto (\velocity, \pressure)$, is nonlinear.

The pressure, $\pressure$, is discretized with first order scalar continuous Galerkin finite elements defined on a mesh of tetrahedra. The velocity, $\velocity$, is discretized with second order continuous Galerkin finite elements on the same mesh. The parameter $\basalfriction$ is discretized with first order scalar continuous Galerkin finite elements on the mesh of triangles that results from restricting the tetrahedral mesh to the basal boundary, $\Gamma_b$. Note that $\Gamma_b$ is a two-dimensional surface embedded in three dimensions due to the mountain topography. The proposed PSF method involves translating impulse responses. Hence it requires either a flat domain, or a notion of local parallel transport. We therefore generate a flattened version of $\Gamma_b$, denoted by $\Omega \subset \mathbb{R}^2$, by ignoring the height coordinate. The parameter $\basalfriction$ is viewed as a function on $\Gamma_b$ for the purpose of solving the Stokes equations, and as a function on $\Omega$ for the purpose of building Hessian approximations and defining the regularization.
%This flattening is necessary because our operator approximation method requires adding points on this surface together to get other points on the surface, and this is only well-defined in flat space. 
The observations are generated by adding multiplicative
Gaussian noise to the tangential component of the velocity field
restricted to the top surface of the geometry. We use $5\%$ noise in all cases, except for Figure~\ref{fig:stokes_reconstructions} and Table \ref{tab:condition_number} where the noise is varied from $1\%$ to $25\%$ and the regularization is determined by the Morozov discrepancy principle for each noise level. The true basal friction
coefficient and resulting velocity fields, which are obtained by
solving~\eqref{Stokeseqn}, are shown in Figure~\ref{fig:true_beta_u}.

@@

\begin{figure}
	\begin{subfigure}{0.32\textwidth}
		\includegraphics[scale=0.17]{mstar_0.25noise_cropped.png}
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
		\includegraphics[scale=0.17]{mstar_0.05noise_cropped.png}
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
	\includegraphics[scale=0.17]{mstar_0.01noise_cropped.png}
    \end{subfigure}
	\caption{(Ice sheet) The log basal friction parameter, with color scale as in Figure~\ref{fig:true_beta}, computed from the PDE constrained optimization problem with noise levels: $25\%$ (left), $5.0\%$ (middle), and $1.0\%$ (right).}
	\label{fig:stokes_reconstructions} 
\end{figure}
